Estimating cardiac contraction through high resolution data assimilation of a personalized mechanical model

### 2017-12-20 chapters/paper2/main.tex
Fix labels and reference in the papers

% \clearpage
\newpage
%% main text
\section{Introduction}
\label{sec:introduction}

@@

\section{Materials and methods}
\label{sec:methods}

\subsection{Data acquisition}
\label{sec:clinical_data}

@@

\subsection{Automated geometry and microstructure creation}
For each patient, a 3D tetrahedral mesh of the LV was
constructed from triangulated segmented surfaces of the endo- and
epicardium corresponding to the frame at the beginning of
atrial systole, Figure \ref{fig:pipeline}. A cut was made at the
ventricular base of the segmentation, so that the mesh cavity volume
and the ultrasound measured volume differed by less than  $1$ ml. Mesh
cells were marked into the 17 AHA regions through the regionally
delineated strain data, and the myocardial fiber orientation, denoted
by $\ef$, were assigned using the algorithm from Bayer et al \cite{bayer2012novel},
with the endo- and epicardial helix fiber angles set to
$\alpha_{\text{endo}} = 60$ and $\alpha_{\text{epi}} = -60$, respectively.

\subsection{Mechanical Model}
We represent the heart as a hyperelastic continuum body, where the coordinates in
the reference ($\Xvec$) and the current ($\xvec$) configuration are
related via the displacement field $\uvec = \xvec-\Xvec$. Furthermore,
we utilize the
deformation gradient, the determinant of the deformation gradient
and, the right Cauchy-Green deformation tensor given by $\F= \I+\nabla
\uvec $,  $J = \det \F$ and $\C = \F^T\F$, respectively. 
% Constitutive relations
To model the passive behavior of the myocardium, the transversely 
isotropic strain energy function proposed in  \cite{holzapfel2009constitutive} is adopted:
\begin{align}
\label{eq:hoa}
  \mathcal{W}(I_1, I_{4\ef}) = \frac{a}{2 b} \left( \exp\left\{ b (I_1 
  - 3)\right\}  -1 \right)
  + \frac{a_f}{2 b_f} \left( \exp\left\{ b_f (I_{4\ef} 
  - 1)_+^2\right\} -1 \right).
\end{align}
Here $I_1 = \tr \C$ and $I_{4\ef}= \ef \cdot (\C \ef)$ are invariants
of $\C$, $(\cdotp)_{+}  = \max\{\cdot, 0\}$, and $a, a_f, b, b_f$ are
material stiffness parameters defining the elastic properties of the myocardium.
% Incompressibility
We follow a common approach and assume that the myocardium is incompressible. 
Incompressibility is incorporated in the model by using a two-field variational 
approach, where we introduce a Lagrange multiplier $p$ which represents the 
hydrostatic pressure, and the term $p(J-1)$ is added to the strain-energy.

@@

For both approaches, the resulting displacement field $\uvec$ and hydrostatic pressure $p$
are determined by using the principle of stationary potential energy
\cite{holzapfel2000nonlinear}, which is based on minimizing the total
energy  $\Pi(\uvec, p)$, which includes internal energy derived from
\eqref{eq:hoa} and external energy. The external energy includes
contributions from the measured cavity pressure $p_{\text{LV}}$, and a
linear spring term at the basal boundary, with spring constant
$k$ = 10.0 kPa. The equilibrium solution is found by solving for the minimum
potential energy, $\delta\Pi(\uvec, p) = 0$.

\subsection{Data Assimilation}
In order to constrain the model to each patient's clinical measurements, we
consider a PDE-constrained optimization problem where
the objective functional is given by the misfit between
simulated and measured strain and volume along with a first order Tikhonov regularization of the
model parameters.
\begin{equation}
  \begin{aligned}
    \label{eq:pde_opt}
    & \underset{m}{\text{minimize}}
    & &  \alpha \left( \frac{V^i - \tilde{V}^i}{V^i} \right)^2
    + (1-\alpha)  \sum_{j= 1}^{17} \sum_{k \in \{c,r,l\}}  \left( \varepsilon_{kj}^i
      -  \tilde{\varepsilon}_{kj}^i \right)^2
    + \lambda \| \nabla m^i \|^2 \\
    & \text{subject to}
    & & \delta \Pi(\uvec, p) = 0.
  \end{aligned}
\end{equation}

Here $V$ and $\varepsilon_{kj}$ are the measured volume and regional Lagrangian strain in
segment $j$ in direction $k$ respectively, and $\tilde{V}^i =
-\frac{1}{3} \int_{\partial \Omega_{\text{endo}}} (\Xvec + \uvec)
\cdotp J\F^{-T}\Nvec  \mathrm{d}S$,  and $\varepsilon_{kj} =
\frac{1}{|\Omega_j|}\int_{\Omega_j}  \mathbf{e}_k^T \nabla \uvec \,
\mathbf{e}_k  \mathrm{d}x.$ 
The parameters $\alpha$ and $\lambda$
control the weights on the different terms, and the sum in the second
term is taken over the seventeen AHA
segments, and the three different strain components (Section \ref{sec:clinical_data}).

The data assimilation procedure is divided into two phases; a passive
and an active phase. For the passive phase we iteratively
  estimate the unloaded configuration and the linear isotropic parameter, $a$ in
\eqref{eq:hoa}, using an algorithm similar to the one described in
\cite{nikou2016effects}, and we set $\alpha= 1.0$,  with $\lambda = 0$ and
$\gamma = 0$, minimizing only the misfit with the measured volumes. The
remaining material parameters are fixed according to [Table 1 row 3 of
\cite{holzapfel2009constitutive}]. For the active phase we fix the material parameter optimized in the
passive phase, choose the control variable $m$ to be $\gamma$ or
$T_a$ for the active strain and active stress model respectively, and
set $\alpha = 0.95$ and $\lambda = 0.01$. This choice of $\alpha$ and
$\lambda$ was based on the analysis done in \cite{balaban}. A summary
of our optimization pipeline is provided to the right in
Figure~\ref{fig:pipeline}.

\begin{figure}[htbp]
\centering
    \includegraphics[width=\textwidth]{models}
\caption{Left: Automated anatomical modeling pipeline to produce AHA
  marked simulation meshes with applied fiber orientations from 3D
  echocardiographic segmentations. Right: Optimization
  pipeline. 1. Unloaded geometry and the linear isotropic
  material parameter $a$ in \eqref{eq:hoa} are estimated iteratively. The unloaded geometry is
  estimated based on the backward displacement method (1a)
  \cite{nikou2016effects} and $a$ is estimated by minimizing the
  difference between simulated and measured volumes (1b). 2. The unloaded
  geometry and the material properties are fixed, and the amount of
  contraction ($\gamma$ for active strain and $T_a$ for active stress)
  is estimated by minimizing the mismatch between simulated and
  measured strain and volume. The active optimization continues to the
  next measurement point until all measurement points in the cycle are covered.}
\label{fig:pipeline}
\end{figure}

\subsection{Implementation details}
We employ a Galerkin finite element method with Taylor-Hood
tetrahedral elements, that is $(\uvec, p) \in \mathbb{P}_2 \times
\mathbb{P}_1 $, with $\mathbb{P}_n$ being the space of piecewise
polynomials of degree $n$.
The solver is implemented in the finite element framework FEniCS
\cite{logg2012automated}, and uses a Newton trust region algorithm
\cite{PETScPackage} to solve nonlinear systems. The minimization of the
model-data misfit functional~\eqref{eq:pde_opt} is accomplished by a sequential
quadratic programming algorithm (SQP)~\cite{kraft1988software}, where
the functional gradient is computed by solving an automatically derived adjoint
equation \cite{farrell2013automated}. In these minimizations an upper
bound of $0.5$ and $500$ kPa is set for the active strain ($\gamma$)
and active stress ($T_a$) control variable respectively, which both
are modeled as functions in $\mathbb{P}_1$, yielding one parameter per
nodal point in the mesh.

\subsection{Contraction analysis}
Although direct physical interpretation of the active
  strain parameter $\gamma$ is difficult, it may be seen as the
  relative shortening of an isolated and unloaded muscle cell. A high
  value of $\gamma$ is therefore an indication of higher contractile
  force in the myocardium, independent of load. We propose that
  the spatially averaged $\gamma$ over the entire LV, denoted by  
  $\overline{\gamma}$, can be used as an index of global
  contractility. Similarly, the active stress parameter $T_a$ is related to force
  development at level of the sarcomeres\cite{goktepe2014generalized},
  and the spatially averaged 
  $T_a$, denoted $\overline{T_a}$ can be used as an index of
  contractility. In addition to the contractility information contained in 
$\overline{\gamma}$ and $\overline{T_a}$, the overall elastic state
of the optimized patient models 
can be used to give estimates of LV elastance. The left ventricular
end-systolic elastance $E\es$, the response of end systolic volume to
increased load, is considered to be one of the major determinants for
cardiac systolic function, and was in \cite{sagawa1977end} proposed
as a global index of ventricular contractility.
It is possible to estimate the end systolic elastance directly 
if the end systolic pressure is known or estimated, 
by perturbing the loading conditions on the optimized model at end
systole while fixing the remaining quatities, and calculating the slope
in the resulting ES pressure-volume curve. More
precisely, if $\plv^{\text{ES}}$ is the end-systolic ventricular
pressure, with cavity volume $V^{\text{ES}}$, 
we change the pressure to $\plv^{\text{ES} + \Delta} =
\plv^{\text{ES}} + \Delta \plv$, resulting in a change in volume,
$V^{\text{ES} +\Delta} = V^{\text{ES}} + \Delta V$. The estimate of end-systolic
elastance can then be calculated by 
\begin{align}
  \tilde{E}\es = \frac{\Delta \plv}{\Delta V}.
  \label{eq:estimate_elastance}
\end{align}

\section{Results and discussion}
\label{sec:results}

\subsection{Matching of strain and volume}
We show the results from two representative simulations in
Figure~\ref{fig:generic_result}, one from the LBBB group and one from
the healthy control group. Snapshots from the calculated unloaded and
end-systolic configurations are depicted. For the unloaded
geometry, we also show the image-based geometry at beginning of atrial
systole, and for the end-systolic configuration we show the
longitudinal strain using both the active stress and active strain
approach. We also show the agreement with the corresponding PV-loops.
% and one of the 51 strain traces over the cardiac cycle shown.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{generic_result}
  \caption{Left: Snap shots of the unloaded geometry in red, and the
    corresponding image based geometry taken at beginning of atrial
    systole in black wire-frame. Middle: Snap shot of end systolic
    configurations using the active strain and active stress approach.
    Color-map shows the end-systolic longitudinal strain.
    Right: Simulated and measured pressure-volume loops for
    these hearts using the active strain and active stress approach.}  
  \label{fig:generic_result}
\end{figure}

@@

In order to visualize the overall match of simulated to measured data,
we show linear regression plots in
Figure~\ref{fig:data_matching}. These results are all based on the
active strain formulation. For
the strain, we separately consider the diastolic and systolic points,
as different types of data were used to  constrain the model in these
two phases, namely volume in the diastolic phase and strain in the
systolic phase. An excellent overall fit was obtained for the optimized
volume ($R^2=1.00$) and systolic strains ($R^2=0.95$). 
Diastolic strains, not used in the optimization, were less well
matched ($R^2=0.31$).

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{data_matching}
  \caption{Scatter plot of simulated ($y$-axis) and measured
    ($x$-axis) strain and volume using the active
    strain approach. Left: Scatter plot of simulated versus
    measured volumes and the best linear least-squares fit of these
    points, (slope = $1.00$). Right: Scatter plot of simulated versus
    measured strain for all segments and all directions, separated
    into the diastole, were only the volume was optimized and systole,
    where both the strain and volume were optimized. For diastole, the slope of
    the best linear fit was $0.31$, while the best linear fit for the
    systolic points had a slope of $0.95$.}
  \label{fig:data_matching}
\end{figure}

@@

Global contraction time courses, $\overline{\gamma}$ and
  $\overline{T}_a$, for each patient were synchronized to the valvular
events to normalize for differing cycle lengths. The average 
and standard deviation of these normalized traces for the
LBBB vs the healthy controls are shown in left of
Figure~\ref{fig:contractility}. The healthy patients had a much higher
level of contraction through the cardiac cycle, and the peak values
were compared using one-way ANOVA, yielding a $P-$value less than
$0.001$ for both the active strain and the active stress approach.

The values of calculated $\tilde{E}\es$ for the healthy and LBBB patients are
shown to the right in Figure~\ref{fig:contractility}. The calculated
elastances of the LBBB group were significantly lower than for the healthy
group, with the comparison between the groups using one-way ANOVA
giving a $P-$value of $0.009$ and $0.003$ for the active strain and
active stress respectively.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{contractility}
  \caption{Extracted biomarkers related to cardiac contractility.
    Left:  Mean value of $T_a$ for the two groups
    synchronized with respect to valvular events (mvc: mitral valve
    closure, avo: aortic valve opening, avc: aortic valve closure,
    mvo: mitral valve opening). Shaded region shows $\pm$ one standard
    deviation.
    Middle:  Mean value of $\gamma$ for the two groups
    synchronized with respect to the same valvular events.
    Right: Estimated values of $\tilde{E}\es$, given by
    \eqref{eq:estimate_elastance} using the active stress and the
    active strain approach. The mean value is depicted for each
    group as a bar, and individual points are also displayed.} 
  \label{fig:contractility}
\end{figure}

>>>>

% \clearpage
\newpage
%% main text
\section{Introduction}
\label{paper2:sec:introduction}

@@

\section{Materials and methods}
\label{paper2:sec:methods}

\subsection{Data acquisition}
\label{paper2:sec:clinical_data}

@@

\subsection{Automated geometry and microstructure creation}
For each patient, a 3D tetrahedral mesh of the LV was
constructed from triangulated segmented surfaces of the endo- and
epicardium corresponding to the frame at the beginning of
atrial systole, Figure \ref{paper2:fig:pipeline}. A cut was made at the
ventricular base of the segmentation, so that the mesh cavity volume
and the ultrasound measured volume differed by less than  $1$ ml. Mesh
cells were marked into the 17 AHA regions through the regionally
delineated strain data, and the myocardial fiber orientation, denoted
by $\ef$, were assigned using the algorithm from Bayer et al \cite{bayer2012novel},
with the endo- and epicardial helix fiber angles set to
$\alpha_{\text{endo}} = 60$ and $\alpha_{\text{epi}} = -60$, respectively.

\subsection{Mechanical Model}
We represent the heart as a hyperelastic continuum body, where the coordinates in
the reference ($\Xvec$) and the current ($\xvec$) configuration are
related via the displacement field $\uvec = \xvec-\Xvec$. Furthermore,
we utilize the
deformation gradient, the determinant of the deformation gradient
and, the right Cauchy-Green deformation tensor given by $\F= \I+\nabla
\uvec $,  $J = \det \F$ and $\C = \F^T\F$, respectively. 
% Constitutive relations
To model the passive behavior of the myocardium, the transversely 
isotropic strain energy function proposed in  \cite{holzapfel2009constitutive} is adopted:
\begin{align}
\label{paper2:eq:hoa}
  \mathcal{W}(I_1, I_{4\ef}) = \frac{a}{2 b} \left( \exp\left\{ b (I_1 
  - 3)\right\}  -1 \right)
  + \frac{a_f}{2 b_f} \left( \exp\left\{ b_f (I_{4\ef} 
  - 1)_+^2\right\} -1 \right).
\end{align}
Here $I_1 = \tr \C$ and $I_{4\ef}= \ef \cdot (\C \ef)$ are invariants
of $\C$, $(\cdotp)_{+}  = \max\{\cdot, 0\}$, and $a, a_f, b, b_f$ are
material stiffness parameters defining the elastic properties of the myocardium.
% Incompressibility
We follow a common approach and assume that the myocardium is incompressible. 
Incompressibility is incorporated in the model by using a two-field variational 
approach, where we introduce a Lagrange multiplier $p$ which represents the 
hydrostatic pressure, and the term $p(J-1)$ is added to the strain-energy.

@@

For both approaches, the resulting displacement field $\uvec$ and hydrostatic pressure $p$
are determined by using the principle of stationary potential energy
\cite{holzapfel2000nonlinear}, which is based on minimizing the total
energy  $\Pi(\uvec, p)$, which includes internal energy derived from
\eqref{paper2:eq:hoa} and external energy. The external energy includes
contributions from the measured cavity pressure $p_{\text{LV}}$, and a
linear spring term at the basal boundary, with spring constant
$k$ = 10.0 kPa. The equilibrium solution is found by solving for the minimum
potential energy, $\delta\Pi(\uvec, p) = 0$.

\subsection{Data Assimilation}
In order to constrain the model to each patient's clinical measurements, we
consider a PDE-constrained optimization problem where
the objective functional is given by the misfit between
simulated and measured strain and volume along with a first order Tikhonov regularization of the
model parameters.
\begin{equation}
  \begin{aligned}
    \label{paper2:eq:pde_opt}
    & \underset{m}{\text{minimize}}
    & &  \alpha \left( \frac{V^i - \tilde{V}^i}{V^i} \right)^2
    + (1-\alpha)  \sum_{j= 1}^{17} \sum_{k \in \{c,r,l\}}  \left( \varepsilon_{kj}^i
      -  \tilde{\varepsilon}_{kj}^i \right)^2
    + \lambda \| \nabla m^i \|^2 \\
    & \text{subject to}
    & & \delta \Pi(\uvec, p) = 0.
  \end{aligned}
\end{equation}

Here $V$ and $\varepsilon_{kj}$ are the measured volume and regional Lagrangian strain in
segment $j$ in direction $k$ respectively, and $\tilde{V}^i =
-\frac{1}{3} \int_{\partial \Omega_{\text{endo}}} (\Xvec + \uvec)
\cdotp J\F^{-T}\Nvec  \mathrm{d}S$,  and $\varepsilon_{kj} =
\frac{1}{|\Omega_j|}\int_{\Omega_j}  \mathbf{e}_k^T \nabla \uvec \,
\mathbf{e}_k  \mathrm{d}x.$ 
The parameters $\alpha$ and $\lambda$
control the weights on the different terms, and the sum in the second
term is taken over the seventeen AHA
segments, and the three different strain components (Section \ref{paper2:sec:clinical_data}).

The data assimilation procedure is divided into two phases; a passive
and an active phase. For the passive phase we iteratively
  estimate the unloaded configuration and the linear isotropic parameter, $a$ in
\eqref{paper2:eq:hoa}, using an algorithm similar to the one described in
\cite{nikou2016effects}, and we set $\alpha= 1.0$,  with $\lambda = 0$ and
$\gamma = 0$, minimizing only the misfit with the measured volumes. The
remaining material parameters are fixed according to [Table 1 row 3 of
\cite{holzapfel2009constitutive}]. For the active phase we fix the material parameter optimized in the
passive phase, choose the control variable $m$ to be $\gamma$ or
$T_a$ for the active strain and active stress model respectively, and
set $\alpha = 0.95$ and $\lambda = 0.01$. This choice of $\alpha$ and
$\lambda$ was based on the analysis done in \cite{balaban}. A summary
of our optimization pipeline is provided to the right in
Figure~\ref{paper2:fig:pipeline}.

\begin{figure}[htbp]
\centering
    \includegraphics[width=\textwidth]{models}
\caption{Left: Automated anatomical modeling pipeline to produce AHA
  marked simulation meshes with applied fiber orientations from 3D
  echocardiographic segmentations. Right: Optimization
  pipeline. 1. Unloaded geometry and the linear isotropic
  material parameter $a$ in \eqref{paper2:eq:hoa} are estimated iteratively. The unloaded geometry is
  estimated based on the backward displacement method (1a)
  \cite{nikou2016effects} and $a$ is estimated by minimizing the
  difference between simulated and measured volumes (1b). 2. The unloaded
  geometry and the material properties are fixed, and the amount of
  contraction ($\gamma$ for active strain and $T_a$ for active stress)
  is estimated by minimizing the mismatch between simulated and
  measured strain and volume. The active optimization continues to the
  next measurement point until all measurement points in the cycle are covered.}
\label{paper2:fig:pipeline}
\end{figure}

\subsection{Implementation details}
We employ a Galerkin finite element method with Taylor-Hood
tetrahedral elements, that is $(\uvec, p) \in \mathbb{P}_2 \times
\mathbb{P}_1 $, with $\mathbb{P}_n$ being the space of piecewise
polynomials of degree $n$.
The solver is implemented in the finite element framework FEniCS
\cite{logg2012automated}, and uses a Newton trust region algorithm
\cite{PETScPackage} to solve nonlinear systems. The minimization of the
model-data misfit functional~\eqref{paper2:eq:pde_opt} is accomplished by a sequential
quadratic programming algorithm (SQP)~\cite{kraft1988software}, where
the functional gradient is computed by solving an automatically derived adjoint
equation \cite{farrell2013automated}. In these minimizations an upper
bound of $0.5$ and $500$ kPa is set for the active strain ($\gamma$)
and active stress ($T_a$) control variable respectively, which both
are modeled as functions in $\mathbb{P}_1$, yielding one parameter per
nodal point in the mesh.

\subsection{Contraction analysis}
Although direct physical interpretation of the active
  strain parameter $\gamma$ is difficult, it may be seen as the
  relative shortening of an isolated and unloaded muscle cell. A high
  value of $\gamma$ is therefore an indication of higher contractile
  force in the myocardium, independent of load. We propose that
  the spatially averaged $\gamma$ over the entire LV, denoted by  
  $\overline{\gamma}$, can be used as an index of global
  contractility. Similarly, the active stress parameter $T_a$ is related to force
  development at level of the sarcomeres\cite{goktepe2014generalized},
  and the spatially averaged 
  $T_a$, denoted $\overline{T_a}$ can be used as an index of
  contractility. In addition to the contractility information contained in 
$\overline{\gamma}$ and $\overline{T_a}$, the overall elastic state
of the optimized patient models 
can be used to give estimates of LV elastance. The left ventricular
end-systolic elastance $E\es$, the response of end systolic volume to
increased load, is considered to be one of the major determinants for
cardiac systolic function, and was in \cite{sagawa1977end} proposed
as a global index of ventricular contractility.
It is possible to estimate the end systolic elastance directly 
if the end systolic pressure is known or estimated, 
by perturbing the loading conditions on the optimized model at end
systole while fixing the remaining quatities, and calculating the slope
in the resulting ES pressure-volume curve. More
precisely, if $\plv^{\text{ES}}$ is the end-systolic ventricular
pressure, with cavity volume $V^{\text{ES}}$, 
we change the pressure to $\plv^{\text{ES} + \Delta} =
\plv^{\text{ES}} + \Delta \plv$, resulting in a change in volume,
$V^{\text{ES} +\Delta} = V^{\text{ES}} + \Delta V$. The estimate of end-systolic
elastance can then be calculated by 
\begin{align}
  \tilde{E}\es = \frac{\Delta \plv}{\Delta V}.
  \label{paper2:eq:estimate_elastance}
\end{align}

\section{Results and discussion}
\label{paper2:sec:results}

\subsection{Matching of strain and volume}
We show the results from two representative simulations in
Figure~\ref{paper2:fig:generic_result}, one from the LBBB group and one from
the healthy control group. Snapshots from the calculated unloaded and
end-systolic configurations are depicted. For the unloaded
geometry, we also show the image-based geometry at beginning of atrial
systole, and for the end-systolic configuration we show the
longitudinal strain using both the active stress and active strain
approach. We also show the agreement with the corresponding PV-loops.
% and one of the 51 strain traces over the cardiac cycle shown.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{generic_result}
  \caption{Left: Snap shots of the unloaded geometry in red, and the
    corresponding image based geometry taken at beginning of atrial
    systole in black wire-frame. Middle: Snap shot of end systolic
    configurations using the active strain and active stress approach.
    Color-map shows the end-systolic longitudinal strain.
    Right: Simulated and measured pressure-volume loops for
    these hearts using the active strain and active stress approach.}  
  \label{paper2:fig:generic_result}
\end{figure}

@@

In order to visualize the overall match of simulated to measured data,
we show linear regression plots in
Figure~\ref{paper2:fig:data_matching}. These results are all based on the
active strain formulation. For
the strain, we separately consider the diastolic and systolic points,
as different types of data were used to  constrain the model in these
two phases, namely volume in the diastolic phase and strain in the
systolic phase. An excellent overall fit was obtained for the optimized
volume ($R^2=1.00$) and systolic strains ($R^2=0.95$). 
Diastolic strains, not used in the optimization, were less well
matched ($R^2=0.31$).

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{data_matching}
  \caption{Scatter plot of simulated ($y$-axis) and measured
    ($x$-axis) strain and volume using the active
    strain approach. Left: Scatter plot of simulated versus
    measured volumes and the best linear least-squares fit of these
    points, (slope = $1.00$). Right: Scatter plot of simulated versus
    measured strain for all segments and all directions, separated
    into the diastole, were only the volume was optimized and systole,
    where both the strain and volume were optimized. For diastole, the slope of
    the best linear fit was $0.31$, while the best linear fit for the
    systolic points had a slope of $0.95$.}
  \label{paper2:fig:data_matching}
\end{figure}

@@

Global contraction time courses, $\overline{\gamma}$ and
  $\overline{T}_a$, for each patient were synchronized to the valvular
events to normalize for differing cycle lengths. The average 
and standard deviation of these normalized traces for the
LBBB vs the healthy controls are shown in left of
Figure~\ref{paper2:fig:contractility}. The healthy patients had a much higher
level of contraction through the cardiac cycle, and the peak values
were compared using one-way ANOVA, yielding a $P-$value less than
$0.001$ for both the active strain and the active stress approach.

The values of calculated $\tilde{E}\es$ for the healthy and LBBB patients are
shown to the right in Figure~\ref{paper2:fig:contractility}. The calculated
elastances of the LBBB group were significantly lower than for the healthy
group, with the comparison between the groups using one-way ANOVA
giving a $P-$value of $0.009$ and $0.003$ for the active strain and
active stress respectively.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{contractility}
  \caption{Extracted biomarkers related to cardiac contractility.
    Left:  Mean value of $T_a$ for the two groups
    synchronized with respect to valvular events (mvc: mitral valve
    closure, avo: aortic valve opening, avc: aortic valve closure,
    mvo: mitral valve opening). Shaded region shows $\pm$ one standard
    deviation.
    Middle:  Mean value of $\gamma$ for the two groups
    synchronized with respect to the same valvular events.
    Right: Estimated values of $\tilde{E}\es$, given by
    \eqref{paper2:eq:estimate_elastance} using the active stress and the
    active strain approach. The mean value is depicted for each
    group as a bar, and individual points are also displayed.} 
  \label{paper2:fig:contractility}
\end{figure}
